---
title: "Prpf8"
output: html_notebook
---

```{r setup}
library(tidyverse)
library(bayesplot)
library(tidybayes)
library(ggdist)
library(brms)
options(mc.cores = 4, brms.backend = "cmdstanr")
theme_set(cowplot::theme_cowplot())
devtools::load_all()
```

```{r}
f <- here::here("private_data/qPCR/qPCR Prpf8 d17 4w final.xlsx")
litter_sheet <- "Litter "
litters <- read_all_litters(f, litter_sheet, c("A2:I10", "A13:I18", "A21:I28", "K2:S8","K11:S21","K24:S34"))
```

```{r}
a1 <- read_qpcr(f, "qPCR set a1", range = "A5:AR31")
a2 <- read_qpcr(f, "qPCR set a2", range = "A5:Z22")
a3 <- read_qpcr(f, "qPCR set a3", range = "A5:AU22")
a4 <- read_qpcr(f, "qPCR set a4", range = "A5:AC56")

qpcr_d17 <- rbind(a1,a2,a3,a4)

```

```{r}
housekeeping <- c("Ubb", "Gapdh", "Actb") 

qpcr_d17 %>% filter(primer %in% housekeeping) %>%
  ggplot(aes(x = Cq)) + geom_histogram(binwidth = 0.2) + facet_grid(genotype ~ primer, scales = "free")
```
```{r}
qpcr_d17 %>% filter(primer %in% housekeeping) %>%
  ggplot(aes(y = Cq, x = genotype)) + geom_boxplot(outlier.shape = NA) + 
  geom_point(aes(color = sex, shape = sex), position = position_jitter(width = 0.25), alpha = 0.7) + 
  facet_wrap( ~ primer, scales = "free")

```

```{r}
qpcr_d17 %>% filter(primer %in% housekeeping) %>%
  group_by(animal_plate, genotype, primer_short, sex) %>%
  summarise(Cq = mean(Cq), .groups = "drop") %>%
  ggplot(aes(y = Cq, x = primer_short, color = sex, group = animal_plate)) + 
  geom_line(alpha = 0.7) + 
  facet_wrap( ~ genotype, scales = "free")

```



```{r, fig.width=10, fig.height=6}
to_compare <- setdiff(unique(qpcr_d17$primer), housekeeping)
step_size <- 6
for(step in 1:ceiling(length(to_compare) / step_size)) {
  primers_to_show <- to_compare[((step - 1) * step_size + 1):(step * step_size)]
  # print(qpcr_d17 %>% filter(primer %in% primers_to_show) %>%
  #   ggplot(aes(x = ct)) + geom_histogram(binwidth = 0.2) + facet_grid(genotype ~ primer, scales = "free_y")
  # )
  print(qpcr_d17 %>% filter(primer %in% primers_to_show) %>%
    ggplot(aes(y = Cq, x = genotype)) + geom_boxplot(outlier.shape = NA) + 
  geom_point(aes(color = sex, shape = sex), position = position_jitter(width = 0.25), alpha = 0.7) + 
  facet_wrap( ~ primer_short, scales = "free")
  )

}

```
```{r}
qpcr_d17 %>% group_by(primer_short, sex) %>% summarise(count = n()) %>% arrange(count)
```


```{r}
#qpcr_d17_model <- qpcr_d17 
 qpcr_d17_model <- qpcr_d17 %>% mutate(primer_short = factor(primer_short),
                                       sex = factor(sex))

priors = c(prior(normal(0,1), class = "sd"),
           prior(normal(0,2), class = "b"))

# fit1 <- brm(Cq ~ primer_short * mo(genotype_ord) + (1 || run) + (primer_short || sex + animal_no + litter + animal_plate + animal_replicate), data = qpcr_d17, prior = priors)
fit1 <- brm(
  bf(Cq ~ primer_short * genotype + sex*primer_short + (1 || run + animal_no + animal_no:primer_short + litter:primer_short),
     sigma ~ (1 || primer_short + run)
     ), 
  data = qpcr_d17_model, 
  prior = priors, refresh = 100,
  file = here::here("local_temp_data/fit1_all"), file_refit = "on_change")
fit1
```
```{r}

nuts_params(fit1) %>% filter(Parameter == "treedepth__") %>% group_by(Value) %>% summarise(n())

```


```{r}
pred_fit1 <- posterior_predict(fit1, cores = 1)
```

```{r}
ppc_stat_grouped(qpcr_d17_model$Cq, pred_fit1, group = qpcr_d17_model$primer_short, stat = "sd")
ppc_stat_grouped(qpcr_d17_model$Cq, pred_fit1, group = qpcr_d17_model$genotype_ord, stat = "sd")
ppc_stat_grouped(qpcr_d17_model$Cq, pred_fit1, group = paste0(qpcr_d17_model$primer_short, qpcr_d17_model$genotype_ord), stat = "sd")
ppc_stat_grouped(qpcr_d17_model$Cq, pred_fit1, group = paste0(qpcr_d17_model$primer_short, qpcr_d17_model$genotype_ord), stat = "mean")
ppc_stat_grouped(qpcr_d17_model$Cq, pred_fit1, group = qpcr_d17_model$sex, stat = "sd")
ppc_stat_grouped(qpcr_d17_model$Cq, pred_fit1, group = qpcr_d17_model$run, stat = "sd")
ppc_stat_grouped(qpcr_d17_model$Cq, pred_fit1, group = qpcr_d17_model$run, stat = "mean")

```

```{r}
qpcr_d17_model %>% group_by(primer_short) %>% summarise(mean = mean(Cq), sd  = sd(Cq)) %>%
  arrange(mean)
```
TODO: Najít a reportovat bug s posterior_epred a mo

```{r}
comparisons <- data.frame(
  numerator = factor(c("MR522/523", "MR512/513", "MR494/495"), levels = levels(qpcr_d17_model$primer_short)),
  denominator = factor(c("MR523/524","MR515/516","MR492/493"), levels = levels(qpcr_d17_model$primer_short)),   
  
  eff_numerator = 1.8, eff_denominator = 1.8)  %>%
  mutate(label = paste0(numerator, " : ", denominator), comparison_id = 1:n())

# pred_data_numerator_wt <- qpcr_d17_model %>% filter(primer_short %in% comparisons$numerator, genotype_ord == "wt",
#                                               sex == "F") %>% 
#   group_by(primer_short) %>% top_n(1)  
# 

pred_data_numerator_wt <- data.frame(primer_short = comparisons$numerator,
                                     genotype = "wt",
                                     sex = "F"
                                     )
pred_data_numerator_d17 <- data.frame(primer_short = comparisons$numerator,
                                     genotype = "d17",
                                     sex = "F"
                                     )
pred_data_denominator_wt <- data.frame(primer_short = comparisons$denominator,
                                     genotype = "wt",
                                     sex = "F"
                                     )
pred_data_denominator_d17 <- data.frame(primer_short = comparisons$denominator,
                                     genotype = "d17",
                                     sex = "F"
                                     )

pred_numerator_wt <- posterior_epred(fit1, newdata = pred_data_numerator_wt, re_formula = NA, cores = 1)
pred_numerator_d17 <- posterior_epred(fit1, newdata = pred_data_numerator_d17, re_formula = NA, cores = 1)
pred_denominator_wt <- posterior_epred(fit1, newdata = pred_data_denominator_wt, re_formula = NA, cores = 1)
pred_denominator_d17 <- posterior_epred(fit1, newdata = pred_data_denominator_d17, re_formula = NA, cores = 1)

pred_numerator <- pred_numerator_wt - pred_numerator_d17 
pred_denominator <- pred_denominator_wt - pred_denominator_d17 

comparisons_pred_numerator <- tidybayes::add_draws(comparisons, pred_numerator, value = "pred_numerator")
comparisons_pred_denominator <- tidybayes::add_draws(comparisons, pred_denominator, value = "pred_denominator")

comparisons_pred <- comparisons_pred_numerator %>% 
  inner_join(comparisons_pred_denominator, by = setdiff(names(comparisons_pred_numerator), "pred_numerator")) %>%
  mutate(log_ratio = log(eff_numerator) * pred_numerator - log(eff_denominator) * pred_denominator)


if(nrow(comparisons_pred) != nrow(comparisons_pred_numerator) || nrow(comparisons_pred) != nrow(comparisons_pred_denominator)) {
  stop("Bad join")
}

```

```{r}

qpcr_d17_averaged <- qpcr_d17_model %>% 
  group_by(genotype, primer_short) %>%
  summarise(avg_Cq = mean(Cq), sd_Cq = sd(Cq), sem_Cq = sd_Cq / n(), .groups = "drop")

qpcr_d17_averaged_comparisons <- comparisons %>%
  inner_join(qpcr_d17_averaged, by = c("numerator" = "primer_short")) %>%
  rename(numerator_avg_Cq = avg_Cq, numerator_sem_Cq = sem_Cq) %>%
  inner_join(qpcr_d17_averaged, by = c("denominator" = "primer_short", "genotype")) %>%
  rename(denominator_avg_Cq = avg_Cq, denominator_sem_Cq = sem_Cq) #%>%
 # mutate(log_ratio = log(eff_numerator) * log(numerator_avg_Cq) - log(eff_denominator) * log(denominator_avg_Cq))

comparisons_obs <- qpcr_d17_averaged_comparisons %>%
  inner_join(qpcr_d17_averaged_comparisons, by = c("label", "comparison_id", "eff_numerator", "eff_denominator"), suffix = c("_d17","_wt")) %>%
  filter(genotype_d17 == "d17", genotype_wt == "wt") %>%
  mutate(log_ratio = log(eff_numerator) * (numerator_avg_Cq_wt - numerator_avg_Cq_d17) -  log(eff_denominator) * (denominator_avg_Cq_wt - denominator_avg_Cq_d17),
         sem_ratio = log(eff_numerator) * (numerator_sem_Cq_wt + numerator_sem_Cq_d17) -  log(eff_denominator) * (denominator_sem_Cq_wt + denominator_sem_Cq_d17))


comparisons_pred %>% ggplot(aes(x = label, y = log_ratio)) + 
  geom_hline(yintercept = 0, color = "blue") + 
  geom_point(data = comparisons_obs, position = position_nudge(x = 0.1), alpha = 0.8) +
  geom_errorbar(data = comparisons_obs, aes(ymin = log_ratio - 1.96 * sem_ratio, ymax = log_ratio + 1.96 * sem_ratio), position = position_nudge(x = 0.1), alpha = 0.8, width = 0.05) +
  stat_pointinterval() + 
  scale_y_continuous("Log[ratio d17/ratio wt]")
```
```{r}
qpcr_d17_replicate_averaged <- qpcr_d17_model %>%
  group_by(run, animal_no, sex, genotype, primer_short) %>%
  summarise(avg_Cq = mean(Cq), .groups = "drop")


ratios_observed <- comparisons %>%
  inner_join(qpcr_d17_replicate_averaged, by = c("numerator" = "primer_short")) %>%
  inner_join(qpcr_d17_replicate_averaged, by = c("denominator" = "primer_short", "run", "animal_no", "sex", "genotype"), suffix = c("_num", "_denom")) %>%
  mutate(Cq_denom_num = avg_Cq_denom - avg_Cq_num)
 
ratios_observed %>%
  ggplot(aes(x = genotype, y = Cq_denom_num)) + geom_boxplot(outlier.shape = NA) + geom_point(aes(color = sex, shape = sex), position = position_jitter(width = 0.3)) + facet_wrap(~label, scales = "free_y") + 
  scale_y_continuous("Cq (denominator) - Cq (numerator)")
```



Steibel et al.: Mixed models, normal error https://www.sciencedirect.com/science/article/pii/S0888754309000986
The same, but Poisson-LogNormal https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0071448



Below, assuming we have "noise free" $Cq$ values.

The ratio of the products $a,b$ in sample $i$ is:

$$
v_i = \frac{E_a^{S_a - Cq_{a,i}}}{E_b^{S_b - Cq_{b,i}}}
$$

Where $E_a, E_b$ are the PCR efficiencies and $S_a, S_b$ - the cycle that would be reached when amplifying a single molecule (similarly an abstract "noise free" value).

We are interested in 

$$
r = \frac{v_1}{v_2} \\
\log{r} = \log v_1 - \log v_2 = \left(  (\log E_a)(S_a - Cq_{a,1}) - (\log E_b)(S_b - Cq_{b,1}) \right) - 
\left(  (\log E_a)(S_a - Cq_{a,2}) - (\log E_b)(S_b - Cq_{b,2}) \right) = \\
(\log E_a)(Cq_{a,2} - Cq_{a,1}) - (\log E_b)(Cq_{b,2}-Cq_{b,1})
$$
